---
title: "RMED901A Exam Report"
author: "Data Drama Queens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read and Tidy


## Tidy, Adjust and Explore

## Plot Creation

We have created plots to answer the following questions:<br />
1. Are there any correlated measurements?<br />
2. Does the white blood cell count distribution depend on pot?<br />
3. Does remission of inflammation after Thiopurines for > 12 weeks change with percent of monocytes in WBC count?<br />

The analysis was conducted in the script [20250909-creating_plots.R]()

### Correlated measurements
The code and resulting plot for the evaluation of correlated measurements are shown below. The plot indicates that some variables are strongly correlated. As an example, we see that `age_days` and `age_years` are strongly correlated. This is reassuring as `age_years` was created as a function of `age_days`. We can also see that mean corpuscular (RBC) hemoglobin (`mch`) and mean corpuscular (RBC) volume (`mcv`), as well as Alanine Transaminase (`alt`) and Aspartate Transaminase (`ast`) are strongly correlated. In addition, we see that the percent of Lymphocytes in WBC count (`lymph_percent`) is strongly negatively correlated with percent of Neutrophils in WBC count (`neut_percent`).

```{r load_libraries_and_file, include=FALSE}
library(ggplot2)
library(tidyverse)
library(here)

patient_data <- read_delim(
  here("data", "data_for_analysis", "20250908-tidy-joined-exam-data.txt"),
  delim = "\t"
)
```

```{r corr_plot, warning=FALSE}
# Create correlation plot
GGally::ggcorr(patient_data,
                           low = "#2166ac",
                           mid = "white",
                           high = "#b2182b",
                           midpoint = 0,
                           label = FALSE,
                           size = 3)

```

### White Blood Cell Count and Potassium


## Statistical Tests
