---
title: "RMED901A Exam Report"
author: "Data Drama Queens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read and Tidy


## Tidy, Adjust and Explore

## Plot Creation

We have created plots to answer the following questions:<br />
1. Are there any correlated measurements?<br />
2. Does the white blood cell count distribution depend on pot?<br />
3. Does remission of inflammation after Thiopurines for > 12 weeks change with percent of monocytes in WBC count?<br />

The analysis was conducted in the script [20250909-creating_plots.R]()

### Correlated measurements
The code and resulting plot for the evaluation of correlated measurements are shown below.

```{r load_libraries_and_file, include=FALSE}
library(ggplot2)
library(tidyverse)
library(here)

patient_data <- read_delim(
  here("data", "data_for_analysis", "20250908-tidy-joined-exam-data.txt"),
  delim = "\t"
)
```

```{r corr_plot}
# Create correlation plot
GGally::ggcorr(patient_data,
                           low = "#2166ac",
                           mid = "white",
                           high = "#b2182b",
                           midpoint = 0,
                           label = FALSE,
                           size = 3)

```



## Statistical Tests
