---
title: "RMED901A Exam Report"
author: "Data Drama Queens"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read and Tidy

## Introduction

This script shows the process of reading raw patient data, exploring its structure, cleaning duplicates, and tidying data by removing or adding columns and saving the cleaned dataset

```{r setup, , include=FALSE}
library(tidyverse)
library(here)
library(skimr)
library(naniar)
```

## Data Reading

To read a text file using `read_delim` and the `here` package for handling file paths, we can use the following script:

```{r Read dataset, echo=TRUE, results='hide'}
patient_data <- read_delim(
here("data", "original_data", "exam_data.txt"),
delim = "\t"
)
```


### Overview of the data

To get a general view of the data structure, we can use the following script:

```{r Data overview, include=TRUE}
glimpse(patient_data) # Check structure and types
head(patient_data) # Show first rows
skimr::skim(patient_data) # Summary statistics
naniar::gg_miss_var(patient_data) # Missing data visualization

ncol(patient_data) # Number of columns
nrow(patient_data) # Number of rows
```

### Renaming Columns

To rename columns to consistent names, use **rename()** function.

```{r column rename, echo=FALSE}
patient_data <- patient_data %>%
rename(
neut_percent = %neut, # Percentage of neutrophils
lymph_percent = lymph%, # Percentage of lymphocytes
age_days = days of life, # Age in days
# Basic Metabolic Panel (BMP) variables
sodium = sod,                     
potassium = pot,                  
chloride = chlor,                 
bicarbonate = co2,                
blood_urea_nitrogen = un          
)
```


### Removing Duplicate Columns

Identify and remove duplicated columns to avoid redundancy by using the following script:

```{r remove duplicates}
patient_data <- patient_data %>%
select(-wbc_copy) # Remove duplicated white blood cell column
```

### Checking Unique Observations
To Count total rows, unique patient IDs, and identify duplicated rows we can use the following script:

```{r Row identification, include=TRUE}
patient_data %>% count() # Total observations
patient_data %>%
distinct(patient_id) %>%
count() # Unique patients
sum(duplicated(patient_data)) # Number of duplicated rows
```

There are 5167 unique patient observations out of 15524 rows, indicating repeated measurements for some patients. Also 23 duplicated rows exist and will be removed.

### Creating New Columns by Pivoting

To tidy the data format, we can create new characteristic columns using `pivot_wider()` and reshape the data so each patient has one row characteristic column.

Here duplicates are removed by keeping distinct rows.

```{r pivoting, include=TRUE}
patient_data <- patient_data %>%
distinct() %>% # Remove duplicated rows
pivot_wider(
names_from = "mean_RBC_characteristic",
values_from = "mean_value"
)
```

### Verify Final Data Structure

Check the dimensions and data types after tidying.

```{r Data verification, include=TRUE}
dim(patient_data) # Expected: 5167 rows, 24 columns
glimpse(patient_data) # Ensure variables are double and data is tidy

```

### Saving Tidied Data

Save the cleaned and tidied dataset as a text file using `write_delim()`.

```{r Data saving, include=TRUE}
filename <- here("data", "data_for_analysis", "20250905-tidy-exam-data.txt")
write_delim(
patient_data,
file = filename,
delim = "\t"
)
```

## Tidy, Adjust and Explore
Following are the codes from the script "20250908-tidy-adjust-explore.data.R". In this script we have futher tidied the data by removing unnecessary columns, joined with additional dataset, created new set of variables and ordered/arranged the variables. 

In this session we have also explored the dataset by stratifying for variables and reporting statistical characteristics of these variables.


### Load the necessary libraries
```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
```


### Read the datasets

We are reading in two dataset (main dataset: `patient_data`, joining dataset: `patient_data_join`) that we will merge at a later point in the script.

```{r read_data, warning=FALSE, message=FALSE}
patient_data <- read_delim(
  here("data", "data_for_analysis", "20250905-tidy-exam-data.txt"),
  delim = "\t"
)

patient_data_join <- read_delim(
  here("data", "original_data", "exam_data_join.txt"),
  delim = "\t"
)
```


### Explore the joining dataset

The joining data appears to have been read in correctly. We see that all the columns/variables in the joining dataset are numeric. 

There are a total of 5167 rows/observations and 10 columns. There are two variables (cal, alt) with missing values. 

There is the same numbers of rows as unique patient IDs which confirms that there are no duplicate IDs.

The joining dataset which will be joined to our main dataset looks ok.

```{r}
glimpse(patient_data_join)
head(patient_data_join)
skimr::skim(patient_data_join)
naniar::gg_miss_var(patient_data_join)

patient_data_join %>%
  count() # Number of rows

patient_data_join %>%
  distinct(patient_id) %>%
  count() # Number of unique patient IDs
```



### Remove unnessary variables abd join the datasets

We remove unnessary columns `hct`and `rdw` and join the two datasets together. 

By looking at the new combined dataset we see that all the variables are still numeric.

We have changed numerical to factor variables as specified in the codebook. 

After looking at the overview of the dataset again, we see that the variables `active` and `remission` are now factors.

```{r}
patient_data <- patient_data %>%
 select(-hct, -rdw) %>% 
 left_join(patient_data_join, join_by("patient_id"))

glimpse(patient_data)

patient_data <- patient_data %>%
  mutate(active = if_else(active == 0, factor("No"), factor("Yes"))) %>%
  mutate(remission = if_else(remission == 0, factor("No"), factor("Yes")))

glimpse(patient_data)
```


### Create new columns

We have created five new columns for later use: `hgb_quartiles`, `blood_ure_nitrogen_over_30`, `lymph_count`, `sodium_fraction` and `age_years`. The dataset have been arranged so that `age_days` and `blood_urea_nitrogen` comes after `patient_id`.

```{r}
patient_data <- patient_data %>% 
  mutate(hgb_quartiles = cut(hgb, breaks=4, labels = c("Q1", "Q2", "Q3", "Q4"))) %>% #Cut the hemoglobin level into quartiles 
  mutate(blood_urea_nitrogen_over_30 = if_else(blood_urea_nitrogen > 30, factor("Yes"), factor("No"))) %>% #create a column indicating if the blood urea nitrogen is above 30
  mutate(lymph_count = wbc * (lymph_percent/100)) %>% #add a column for Lymphocytes cell count
  mutate(sodium_fraction = (sodium / (sodium + potassium + chloride))) %>% #sodium as a fraction of summed sodium, potassium, and chloride
  select(patient_id, age_days, blood_urea_nitrogen, everything()) %>% # Set the order of columns
  arrange(patient_id) # arrange by patient ID

patient_data <- patient_data %>%
  mutate(age_years = round(age_days / 365.25)) # New variable for age in years
```


### Explore the new dataset after joining with new columns

We have counted the categorical variables to confirm that they were created correctly. 

We see from glimpsing the dataset that the new numerical variables seem to be created correctly. 

There are now 4 factor and 32 numerical variables.

There is most missing data for blood_urea_nitrogen and the column based on this (blood_urea_nitrogen_over_30). There is also missing values for 15 other columns.

```{r}
patient_data %>%
  count(hgb_quartiles) # Verify that the new column for hemoglobin quartiles makes sense

patient_data %>%
  count(blood_urea_nitrogen_over_30) # Verify that the categorical column for blood urea nitrogen makes sense


glimpse(patient_data)
skimr::skim(patient_data)
naniar::gg_miss_var(patient_data)
```


### Stratification and report min, max, mean and sd of variables

Here we answer the following:
1.Stratify the data by `hgb_quartiles` and report min, max, mean and sd of a `rbc`.
2.Stratify the by a `hgb_quatiles` and report min, max, mean and sd of a numeric column for a defined set of observations:
+Only for persons with hgb <= 10
+Only for persons with remission of inflamation
+Only for persons older than around 40 years of age
+Only for persons with more than 10% of monocytes in WBC
3.Use two categorical columns `active` and `remission` in the dataset to create a table

```{r}
# Stratification by hgb_quartiles and report min, max, mean and sd of rbc (red blood cell counts)
patient_data %>% 
  group_by(hgb_quartiles) %>%
  summarise(
    min_rbc = min(rbc, na.rm = TRUE),
    max_rbc = max(rbc, na.rm = TRUE),
    mean_rbc = mean(rbc, na.rm = TRUE),
    sd_rbc = sd(rbc, na.rm = TRUE)
  )

# among patients with hgb less than 10
patient_data %>%
  filter(hgb <= 10) %>%  
  group_by(hgb_quartiles) %>%          
  summarise(
    min_rbc_hgb10 = min(rbc, na.rm = TRUE),
    max_rbc_hgb10 = max(rbc, na.rm = TRUE),
    mean_rbc_hgb10 = mean(rbc, na.rm = TRUE),
    sd_rbc_hgb10 = sd(rbc, na.rm = TRUE)
  )

# among patients with remission of inflammation
patient_data %>% 
  filter(remission == "Yes") %>%
  group_by(hgb_quartiles) %>%
  summarise(
    min_rbc_remission = min(rbc, na.rm = TRUE),
    max_rbc_remission = max(rbc, na.rm = TRUE),
    mean_rbc_remission = mean(rbc, na.rm = TRUE),
    sd_rbc_remission = sd(rbc, na.rm = TRUE)
  )

# among patients older than around 40 years of age
patient_data %>%
  group_by(hgb_quartiles) %>%
  filter(age_years > 40) %>% # Already created a variable for age in years earlier
  summarize(mean_rbc_age = mean(rbc, na.rm = TRUE),
            min_rbc_age = min(rbc, na.rm = TRUE),
            max_rbc_age = max(rbc, na.rm = TRUE),
            sd_rbc_age = sd(rbc, na.rm = TRUE))

# among patients with more than 10% of monocytes in WBC
patient_data %>%
  group_by(hgb_quartiles) %>%
  filter(mono_percent > 10) %>%
  summarize(mean_rbc_mono = mean(rbc, na.rm = TRUE),
            min_rbc_mono = min(rbc, na.rm = TRUE),
            max_rbc_mono = max(rbc, na.rm = TRUE),
            sd_rbc_mono = sd(rbc, na.rm = TRUE))

# Create a table of remission vs active inflammation
table(patient_data$remission, patient_data$active)
```


### Saving the joined dataset
```{r}
# Save the joined data ----
filename <- paste0(here("data", "data_for_analysis", "20250908-tidy-joined-exam-data.txt"))
write_delim(
  patient_data, 
  file = filename,
  delim = "\t"
)
```


## Creating Plots

We have created plots to answer the following questions:<br />
1. Are there any correlated measurements?<br />
2. Does the white blood cell count distribution depend on pot?<br />
3. Does remission of inflammation after Thiopurines for > 12 weeks change with percent of monocytes in WBC count?<br />

The analysis was conducted in the script [20250909-creating_plots.R]()

### Correlated measurements
First, we evaluated the presence of correlated measurements in our dataset. The code and resulting plot are shown below. The plot indicates that some variables are strongly correlated. As an example, we see that `age_days` and `age_years` are strongly correlated. This is reassuring as `age_years` was created as a function of `age_days`. We can also see that mean corpuscular (RBC) hemoglobin (`mch`) and mean corpuscular (RBC) volume (`mcv`), as well as Alanine Transaminase (`alt`) and Aspartate Transaminase (`ast`) are strongly correlated. In addition, we see that the percent of Lymphocytes in WBC count (`lymph_percent`) is strongly negatively correlated with percent of Neutrophils in WBC count (`neut_percent`).

```{r load_libraries_and_file, include=FALSE}
library(ggplot2)
library(tidyverse)
library(here)

patient_data <- read_delim(
  here("data", "data_for_analysis", "20250908-tidy-joined-exam-data.txt"),
  delim = "\t"
)
```

```{r corr_plot, warning=FALSE}
# Create correlation plot
GGally::ggcorr(patient_data,
                           low = "#2166ac",
                           mid = "white",
                           high = "#b2182b",
                           midpoint = 0,
                           label = FALSE,
                           size = 3)

```

### White Blood Cell Count and Potassium
Next, we evaluated if the white blood cell count distribution depends on `potassium`. The code and resulting plot are provided below. From the plot, the white blood cell count does not seem to change with potassium level.

```{r wbc_pot, warning=FALSE, message=FALSE}
# Scatterplot investigating white blood cell count by potassium level ----
ggplot(patient_data, aes(x = wbc, y = potassium)) + # Assign data, x and y values
  geom_point(colour = "darkblue", alpha = 0.5, na.rm = TRUE) + # Scatterplot
  geom_smooth(method = "lm", se = FALSE, colour = "orange") + #Regression line
  theme_minimal() + #Background theme
  theme(legend.position = "none", #Remove legend
        plot.title = element_text(face = "bold", size = 12, margin = margin(10, 0, 10, 0)), #Bold title, resized title and increased margin between title and plot
        axis.title.x = element_text(face = "bold", size = 10), #Bold x-axis text
        axis.title.y = element_text(face = "bold", size = 10)) + #Bold y-axis text
  labs(title = "White Blood Cell Count by Potassium Level", #Title
       x = "White blood cell count (10,000 cells per microliter)", #x-axis title
       y = "Potassium level (mmol/L)") #y-axis title

```

### Remission and Monocytes
Lastly, we evaluated if remission of inflammation after Thiopurines for > 12 weeks change with percent of monocytes in WBC count. The code and resulting plot are shown below. From a visual inspection of the boxplot, there seems to be no difference in percent of monocytes in WBC count for groups with or without remission of inflammation after Thiopurines for > 12 weeks. 

```{r}
# Remission and percent of monocytes ----
# Create a boxplot
ggplot(data=patient_data, 
                              mapping=aes(x=remission, y=mono_percent)) + #select variables
  geom_boxplot(fill=c("#7876B199", "#20854E99")) + #create a box plot and set specific colours
  xlab("Remission") + #change x axis label
  ylab("Monocytes in WBC count (%)") + #y axis labels
  labs(title = "Percent of monocytes in WBC count",
       subtitle = "Grouped by remission status") +
  theme_minimal() + #use the minimal theme
  theme(
    panel.grid.major = element_blank(), #remove major grid
    panel.grid.minor = element_blank(), #remove minor grid
    axis.line = element_line(colour="grey", linewidth = 0.5), #include x and y axis
    axis.ticks = element_line(colour = "grey", linewidth = 0.5), #include tick marks
    legend.position = "None", #remove legend
    plot.title = element_text(size = 16, face = "bold"), #change title font
    axis.title = element_text(size = 14), #change axis title size
    axis.text.x = element_text( size = 12), #change x-axis text size
    axis.text.y = element_text(size=12) #change y-axis text size
  ) 
  
 

```





## Statistical Tests
